Lack of Evidence?: Academic Probation Does Not Affect Student Dropout and Subsequent Performance

### 2021-04-22 outputs/paper/paper.Rmd
Commit for README and Discussion

---
title: "Lack of Evidence?: Academic Probation Does Not Affect Student Dropout and Subsequent Performance"
subtitle: "Revisiting the Ability, Gender, and Performance Standards: Evidence from Academic Probation (Lindo, Sanders, and Oreopoulos 2010)"
author: "Hong Shi"
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: " Many universities use academic probation as a wake-up call to ensure currently enrolled students to achieve minimum academic standards. This paper examines the causal impact on first-year students' responses in a Canadian univesity after being placed on academic probation in a sharp regression discontinuity design. The results do not suggest that placing first year students on academic probation affects the dropout decision, subsequent annual GPA and graduation. While there are heterogeneous responses across students' highschool performance, gender, first language towards academic probation. #High Level Importance \\par\\textbf{Keywords:} Regression Discontinuity Design, Bandwidth, Academic Probation, Student Performance, P-value, Causality"
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
toc: FALSE

@@

library(tidyverse)
library(rdrobust)
library(here)
library(stargazer)
library(kableExtra)
library(finalfit)
library(broom)
#install.packages("rdpower")
library(rdpower)
library(modelsummary)
```

@@

&nbsp; &nbsp; This paper examines the causal impact on first-year university students after being placed on academic probation by exploiting the discontinuous probation GPA cutoff in a sharp regression discontinuity design (RDD). The data comes form a large Canadian university with three individual campuses. It is a replication of `“Ability, Gender, and Performance Standards: Evidence from Academic Probation."` [@cite_original] with adjustment on the bandwidth (i.e., student within certain grade points of the academic probation cutoff). After applying the adjusted bandwidth, the results are NOT consistent with Lindo's findings that being placed on probation at the end of the first year increases the student dropout, improves the GPAs of the remaining students and negatively affects graduation rates. While students with different gender, highschool academic performance or first language react differently to academic probation. Furthermore, I also find heterogeneous responses to academic probation across campus that is not mentioned in the original paper (Note: May adjust the result after running the model.).

&nbsp; &nbsp; The rest of this paper is organized as follows: In Data section (Section \@ref(Data)), I would first explain the domain background about the academic probation. Next, I introduce the data used in the original paper. And then, I would describe the Quasi-experiment design rationale and also raise concerns about the bandwidth used in the original paper. In Model section (Section \@ref(mainmodel)), I elaborate on the sharp RDD, address its assumptions and propose the model used in this study. In Results section (Section \@ref(result)), I show the impact of academic probation on student dropout, subsequent performance and graduation with the adjusted bandwidth. In Discussion section (Section \@ref(discuss)), I would first summarize the study findings, and then elaborate on the causality of the experiment, address external and internal validity and also Simpson's Paradox (also possibly Berkson's Paradox) regarding the campus variation. Finally, I conclude with some limitations and suggest future directions of this experiment. ^[Code and data are available at the GitHub repo: https://github.com/honn-ishinn/RDD_academic_probation.]

@@

&nbsp; &nbsp; The data comes from a large Canadian university with three individual campuses: one central campus (Campus 1) and two smaller satellite campuses (Campus 2 and Campus 3) from 1996 to 2005. I analyzed it using R [@citeR], and packages `tidyverse` [@cite_tidyverse],  `here` [@cite_here], `rdrobust` [@cite_rdrobust], `rdd`[@cite_rdd], `rdpower`[@cite_rdpower], `haven`[@cite_haven] . I used packages `bookdown` [@cite_bookdown], `kableExtra` [@cite_kableextra], , `finalfit` [@cite_finalfit], `modelsummary`[@cite_modelsummary], `broom` [@cite_broom] to format the document and referenced `Impact Evaluation in Practice` [@cite_impact] to evaluate this experiment.

&nbsp; &nbsp; The principle of academic probation in this university is simple: if a student's grade point average (GPA) is below a certain threshold, the student is placed on academic probation. At Campus 1 and Campus 2, students with a cumulative GPA below 1.5 grade points are placed on academic probation. Campus 3 has a GPA cutoff at 1.6 grade points. Students with GPA exactly at the probation cutoff are considered in good academic standing.

&nbsp; &nbsp; Students on academic probation will be suspended in subsequent sessions if their grades do not improve. At all campuses, students on probation can avoid suspension and return to good academic standing by improving their cumulative GPA up to the cutoff. Students who fail to sufficiently improve their grades are suspended for one full academic year. If suspended students return to the university and again fail to sufficiently improve their grades, they can be suspended for three years. A third failure to meet the GPA requirement can lead to permanent suspension from all campuses. Students will be notified after being placed on academic probation. A sample letter sent to students at Campus 2 is in Appendix 1.

4. Restrict students having their academic standing evaluated at the end of Year 1 ($98\%$ of the remaining sample).
```{r, include=FALSE}
clean_data <- read.csv(here("inputs/data/cleaned_original.csv"))
```

@@

@@

&nbsp; &nbsp; Instead of a randomized experiment, this study belongs to the Quasi-experiment because students' academic statues are not randomly assigned by the university administration. As the probation cutoff score is clearly defined and the GPAs affect students' academic statuses, the regression discontinuity design (RDD) is the method used to evaluate the probation impact. Therefore, the cutoff rule assigns students barely below the probation cutoff (On probation) into the treatment group and those barely above the probation cutoff (In good standing) into the comparison group to estimate impact on first-year university students after being placed on academic probation.

@@

&nbsp; &nbsp; Since the probation cutoff GPA lies between "adequate" C- of 1.7 GPA and "marginal" D+ of 1.3 GPA, I adjust the bandwidth to be 0.2 to reduce the academic performance difference. Table \@ref(tab:mycutoff) shows the summary statistics of 4106 student observable characteristics within 0.2 GPA. Even though the t-test (Table \@ref(tab:ttesthsme)) on highschool grade percentile still suggests difference on highschool performance, the estimated highschool grade percentile difference is much less than those within 0.6 bandwidth. Besides, the t-test (Table \@ref(tab:ttestcrme)) on course credit does not show a significant difference in course load within 0.2 bandwidth, so the observed characteristics become more balanced.

@@

@@

## Test on Regression Discontinuity Design Assumptions

@@

&nbsp; &nbsp; Since the academic probation cutoff rule is known, one may question that students barely below the probation cutoff would study harder to avoid being placed on academic probation. However, the survey conducted in Lindo's paper suggests that first year students are generally less familiar with university policies to know the GPA required to avoid academic probation. And Lindo also finds no evidence that students' grades are related to their knowledge about academic probation [@cite_original]. So first year students are not likely to "play with" the university policy to avoid academic probation. The following figure (Figure \@ref(fig:histcutoff)) shows the distribution of the GPA distance from probation cutoff of all students. There is a smooth increase on the frequency distribution of the student GPA until reaching students' average GPA (GPA distance from probation around 1) and there is no obvious jump on either side of the probation cutoff. So students on either side of the probation cutoff is balanced as suggested in Data Section \@ref(bandwidth).

@@

ggplot(clean_for_summary, aes(x = dist_from_cut))+
  geom_histogram(aes(y  = stat(count)), binwidth = 0.125, color = "black", fill = "#33CCCC")+
  geom_vline(xintercept = 0, linetype = "dotted")+
  theme_minimal()+
  theme(plot.title = element_text(hjust = 0.5)) +
  ylab("Frequency Headcount")+
  xlab("Year 1 GPA minus Probation Cutoff")+
  ggtitle("Distribution of Year 1 GPA relative to Probation Cutoff for All Students")

@@

&nbsp; &nbsp; By placing students on academic probation, the university informs these students an urge to improve their academic performance to prevent further escalating penalties such as suspension. According to ${B}\acute{e}{nabou}$ and Tirole's model of agents' responses to a performance standard [@cite_model_background], the probation policy as a negative incentive may result in the following outcomes:

@@

$$Y_{i} = \beta_{0}Proation_{i} + f({Cutoff}\:{Distance}_{i}) + \alpha_{1}Gender_{i}+\alpha_{2}HighSchool_{i}+\alpha_{3}Language_{i} +\epsilon_{i}$$

@@

* $Probation_{i}$ is an indicator variable to show whether the student $i$ is placed on academic probation or not in the first year. And its coefficient $\beta_{0}$ is the estimate of impact on first year students after being placed on academic probation that we are primarily interested in.

* $f({Cutoff}\:{Distance}_{i})$ is the function to represent the student $i$'s first year GPA distance from the cutoff. Since the first order polynomial is used during the power analysis to determine required student samples (Data Section \@ref(bandwidth)), it will also be applied in building the sharp RDD model^[Adjusting the order of distance from cutoff function does not affect the results. Lindo's original paper also used first order polynomial on distance from the cutoff to estimate probation outcomes].

* $Gender_{i}$ is an indicator variable to show whether the student $i$ is male or female. Its coefficient $\alpha_{1}$ is the estimate of gender difference in response to academic probation.

* $HighSchool_{i}$ is an indicator variable to show whether the student $i$ has a satisfactory highschool performance (Highschool Grade Percentile $>$ 50) or a relatively poor highschool performance (Highschool Grade Percentile $\leq$ 50). Its coefficient $\alpha_{2}$ is the estimate of highschool performance difference in response to academic probation.

* $Language_{i}$ is an indicator variable to show whether the student $i$'s first language is English or other language. Its coefficient $\alpha_{3}$ is the estimate of first language difference in response to academic probation.

@@

&nbsp; &nbsp; As mentioned in Model Section \@ref(model), logistic regression is used to estimate the probation impact on probability of student dropout.
$$Prob(Dropout_{i}) = \beta_{0}Proation_{i} + f({Cutoff}\:{Distance}_{i}) + \alpha_{1}Gender_{i}+\alpha_{2}HighSchool_{i}+\alpha_{3}Language_{i} +\epsilon_{i}$$

@@

&nbsp; &nbsp; As mentioned in Model Section \@ref(model), linear regression is used to estimate the probation impact on student second year annual GPA.
$$GPAYear2_{i} = \beta_{0}Proation_{i} + f({Cutoff}\:{Distance}_{i}) + \alpha_{1}Gender_{i}+\alpha_{2}HighSchool_{i}+\alpha_{3}Language_{i} +\epsilon_{i}$$

@@

&nbsp; &nbsp; Within 0.2 GPA bandwidth, there is significant difference on student second year GPA across gender and highschool performance. Compared to female students, male students earn 0.120 less annual GPA in the second year, holding other baseline characteristics constant. Compared to students with relative poor highschool academic performance, students with satisfactory academic performance earn 0.099 more annual GPA in the second year, holding other baseline characteristics constant. No other significant difference is found on student first language.
```{r, include=FALSE}
# Linear Regression to estimate the impact of probation on student Year 2 GPA

@@

&nbsp; &nbsp; As mentioned in Model Section \@ref(model), logistic regression is used to estimate the probation impact on probability of student graduating by Year 4.
$$Prob(Graduation_{i}) = \beta_{0}Proation_{i} + f({Cutoff}\:{Distance}_{i}) + \alpha_{1}Gender_{i}+\alpha_{2}HighSchool_{i}+\alpha_{3}Language_{i} +\epsilon_{i}$$
&nbsp; &nbsp; The model baseline characteristics are the same as mentioned in dropout Section \@ref(baseline). As shown in the following table (Table \@ref(tab:gradsummary)), within 0.2 GPA bandwidth, compared to students on academic probation, students in good standing have $26.8\%$ increase in the odds of graduating by Year 4, while holding other baseline characteristics constant. **However**, the p-value of 0.161 suggests that the graduation does not significantly differ between good standing students and probation students, holding other baseline characteristics constant. And this difference could also not be clearly detected around the probation cutoff as shown(Figure \@ref(fig:gradfig)). Within 0.3 GPA and 0.6 GPA bandwidth, the results also do not suggest difference on graduation between good standing students and probation students at 0.05 significance level, holding other baseline characteristics constant.

@@

&nbsp; &nbsp; According to the student summary statistics (Table \@ref(tab:lindo)) and t-test conducted on students highschool performance (Table \@ref(tab:ttesthslin)) and credit attempted in the first year (Table \@ref(tab:ttestcrlin)) in Data Section \@ref(bandwidth), within the bandwidth of 0.6 GPA, students barely below the cutoff generally hold weaker academic potentials than those barely above the cutoff. And these academic differences also serves as key factors affecting student outcomes.

@@

```{r, include=FALSE}
library(scales)
  ggplot(clean_cut,aes(x = study_campus, fill = status))+
  geom_bar(aes(y = (..count..)/sum(..count..)), position = position_dodge())+
  scale_y_continuous(labels = scales::percent)+
  theme_minimal()
```

## Causal inference

&nbsp; &nbsp; So does placing first year students on academic probation affect student outcomes? By applying a narrower bandwidth of 0.2 GPA, student characteristics become more balanced (Table \@ref(tab:mycutoff)). Although the t-test still suggests a difference on highschool performance between treatment and comparison students (Table \@ref(tab:ttesthslin)), model results in Result Section \@ref(result) has already suggestted

>>>>

---
title: "Lack of Evidence?: Academic Probation Does Not Affect Student Dropout and Subsequent Performance"
subtitle: "Revisiting the Ability, Gender, and Performance Standards: Evidence from Academic Probation (Lindo, Sanders, and Oreopoulos 2010)"
author: "Hong Shi"
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: "Many universities use academic probation as a wake-up call to ensure currently enrolled students to achieve minimum academic standards. This paper examines the causal impact on first-year students' responses in a Canadian university after being placed on academic probation in a sharp regression discontinuity design. With adjusted bandwidth, the results do not suggest that placing first year students on academic probation affects the dropout decision, subsequent annual GPA and graduation. While there are heterogeneous responses across students' highschool performance, gender, first language towards academic probation. #High Level Importance \\par\\textbf{Keywords:} Regression Discontinuity Design, Bandwidth, Academic Probation, Student Performance, P-value, Causality"
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
toc: FALSE

@@

library(tidyverse)
library(rdrobust)
library(here)
library(stargazer)
library(kableExtra)
library(finalfit)
library(broom)
#install.packages("rdpower")
library(rdpower)
library(modelsummary)
#install.packages("ggpubr")
library(ggpubr)
```

@@

&nbsp; &nbsp; This paper examines the causal impact on first-year university students after being placed on academic probation by exploiting the discontinuous probation GPA cutoff in a sharp regression discontinuity design (RDD). The data comes form a large Canadian university with three individual campuses. It aims to replicate main findings in `“Ability, Gender, and Performance Standards: Evidence from Academic Probation."`^[Original paper available at: https://www.aeaweb.org/articles?id=10.1257/app.2.2.95] [@cite_original] with adjustment on the bandwidth (i.e., student within certain grade point average of the academic probation cutoff). After applying the adjusted bandwidth, the results are NOT consistent with Lindo's findings that being placed on probation at the end of the first year increases the student dropout, improves the GPAs of the remaining students and negatively affects graduation rates. While students with different gender, highschool academic performance or first language react differently to academic probation. Furthermore, I also find heterogeneous responses towards academic probation across campus that is not elaborated in the original paper.

&nbsp; &nbsp; The rest of this paper is organized as follows: In Data section (Section \@ref(Data)), I would first explain the domain background about the academic probation. Next, I introduce the data used in the original paper. And then, I would describe the Quasi-experiment design rationale and also raise concerns about the bandwidth used in the original paper. In Model section (Section \@ref(mainmodel)), I elaborate on the sharp RDD, address its assumptions and propose the model used in this study. In Results section (Section \@ref(result)), I show the impact of academic probation on student dropout and subsequent academic performance with the adjusted bandwidth. In Discussion section (Section \@ref(discuss)), I would first comment on the study results, and then elaborate on the causality of this quasi-experiment, address internal and external validity regarding the campus variation. I would also discuss potential bias in this experiment. Finally, I conclude with some weakness and suggest future directions of this study. ^[Code and data are available at the GitHub repo: https://github.com/honn-ishinn/RDD_academic_probation.]

@@

&nbsp; &nbsp; The data comes from a large Canadian university with three individual campuses: one central campus (Campus 1) and two smaller satellite campuses (Campus 2 and Campus 3) from 1996 to 2005. I analyzed it using R [@citeR], and packages `tidyverse` [@cite_tidyverse],  `here` [@cite_here], `rdrobust` [@cite_rdrobust], `rdd`[@cite_rdd], `rdpower`[@cite_rdpower], `haven`[@cite_haven] . I used packages `bookdown` [@cite_bookdown], `kableExtra` [@cite_kableextra], , `finalfit` [@cite_finalfit], `modelsummary`[@cite_modelsummary], `broom` [@cite_broom], `ggpubr` [@cite_ggpubr] to format the document and referenced `Impact Evaluation in Practice` [@cite_impact] to evaluate this experiment.
```{r, include=FALSE}
clean_data <- read.csv(here("inputs/data/cleaned_original.csv"))
```

@@

&nbsp; &nbsp; The principle of academic probation in this university is simple: if a student's grade point average (GPA) is below a certain threshold, the student is placed on academic probation. At Campus 1 and Campus 2, students with a cumulative GPA below 1.5 grade points are placed on academic probation. Campus 3 has a GPA cutoff at 1.6 grade points. Students with GPA exactly at the probation cutoff are considered in good academic standing. The following histograms on student highschool grade and attempted course credit in first year suggest that students on academic probation generally have lower highschool grades (Figure \@ref(fig:histhsgrade)), and attempt less courses for credits in the first year(Figure \@ref(fig:histcredit)).

```{r, include=FALSE}
#Filter invalid data caused by university administrative error in data reporting

clean_summary_addcol <- clean_for_summary %>% 
  filter((dist_from_cut < 0 & probation_year1 ==1)| (dist_from_cut >=0 & probation_year1 ==0)) %>% 
  mutate( status = case_when(probation_year1 == 0 ~ "In Good Standing",
                             probation_year1 == 1 ~ "On Academic Probation"))

# Set students on academic probation as the baseline group for better visualization
clean_summary_addcol$status <- as.factor(clean_summary_addcol$status)
clean_summary_addcol$status <- relevel(clean_summary_addcol$status, "On Academic Probation")
```

```{r histhsgrade, fig.align='center', fig.cap="Distribution of Highschool Grade by Academic Status for All Students", fig.width=8, fig.height=4, echo= FALSE, message=FALSE, warning=FALSE}
# Referenced the following to plot the histogram
# http://www.sthda.com/english/wiki/ggplot2-histogram-plot-quick-start-guide-r-software-and-data-visualization
# Histogram of the distribution of GPA from cutoff

ggplot(clean_summary_addcol, aes(x = hsgrade_pct, color = status, fill = status))+
  geom_histogram(aes(y  = ..density..), binwidth = 2.5, position = "identity", alpha = 0.4)+
  theme_minimal()+
  theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom", legend.title = element_blank()) +
  ylab("Density")+
  xlab("Highschool Grade Percentile")+
  ggtitle("Distribution of Highschool Grade by Academic Status for All Students")
```

```{r histcredit, fig.align='center', fig.cap="Distribution of Attempted Credit in Year 1 by Academic Status for All Students", fig.width=8, fig.height=4, echo= FALSE, message=FALSE, warning=FALSE}
# Histogram of the distribution of attempted credit in Year 1

ggplot(clean_summary_addcol, aes(x = totcredits_year1, color = status, fill = status))+
  geom_histogram(aes(y  = ..density..), binwidth = 0.5, position = "identity", alpha = 0.4)+
  theme_minimal()+
  theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom", legend.title = element_blank()) +
  ylab("Density")+
  xlab("Highschool Grade Percentile")+
  ggtitle("Distribution of Attempted Credit in Year 1 by Academic Status for All Students")
```

&nbsp; &nbsp; Students on academic probation will be suspended in subsequent sessions if their grades do not improve. At all campuses, students on probation can avoid suspension and return to good academic standing by improving their cumulative GPA up to the cutoff. Students who fail to sufficiently improve their grades are suspended for one full academic year. If suspended students return to the university and again fail to sufficiently improve their grades, they can be suspended for three years. A third failure to meet the GPA requirement can lead to permanent suspension from all campuses. Students will be notified after being placed on academic probation. A sample letter sent to students at Campus 2 is in Appendix (Appendix 1 \@ref(letter)).

4. Restrict students having their academic standing evaluated at the end of Year 1 ($98\%$ of the remaining sample).

@@

&nbsp; &nbsp; Instead of a randomized experiment, this study belongs to the quasi-experiment because students' academic statues are not randomly assigned by the university administration. As the probation cutoff score is clearly defined and the GPAs affect students' academic statuses, the regression discontinuity design (RDD) is the method used to evaluate the probation impact. Therefore, the cutoff rule assigns students barely below the probation cutoff (On probation) into the treatment group and those barely above the probation cutoff (In good standing) into the comparison group to estimate impact on first-year university students after being placed on academic probation.

@@

&nbsp; &nbsp; Since the probation cutoff GPA lies between "adequate" C- of 1.7 GPA and "marginal" D+ of 1.3 GPA, I adjust the bandwidth to be 0.2 to reduce the academic performance difference. Table \@ref(tab:mycutoff) shows the summary statistics of 4106 student observable characteristics within 0.2 GPA. Even though the t-test (Table \@ref(tab:ttesthsme)) on highschool grade percentile still suggests difference on highschool performance, the estimated highschool grade percentile difference is much less than those within 0.6 bandwidth. Many probation students with low highschool grade are excluded after applying the narrower 0.2 GPA bandwidth as shown in the following histogram (Figure \@ref(fig:combinehs)). Besides, the t-test (Table \@ref(tab:ttestcrme)) on course credit does not show a significant difference in course load within 0.2 bandwidth, so the observed characteristics become more balanced.

@@

```{r, include=FALSE}
# Set students on academic probation as the baseline group for easier visualizations
clean_cut$status <- as.factor(clean_cut$status)
clean_cut$status <- relevel(clean_cut$status, "On Academic Probation")
clean_cut2$status <- as.factor(clean_cut2$status)
clean_cut2$status <- relevel(clean_cut2$status, "On Academic Probation")
```

```{r, include=FALSE}
# Distribution of Highschool Grades within 0.6 GPA bandwidth
hs_06 <- ggplot(clean_cut2, aes(x = hsgrade_pct, color = status, fill = status))+
  geom_histogram(aes(y  = ..density..), binwidth = 2.5, position = "identity", alpha = 0.4)+
  theme_minimal()+
  theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom", legend.title = element_blank()) +
  ylab("Density")+
  xlab("Highschool Grade Percentile")+
  ggtitle("Within 0.6 GPA Bandwidth")
```

```{r, include=FALSE}
# Distribution of Highschool Grades within 0.2 GPA bandwidth
hs_02 <- ggplot(clean_cut, aes(x = hsgrade_pct, color = status, fill = status))+
  geom_histogram(aes(y  = ..density..), binwidth = 2.5, position = "identity", alpha = 0.4)+
  theme_minimal()+
  theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom", legend.title = element_blank()) +
  ylab("Density")+
  xlab("Highschool Grade Percentile")+
  ggtitle("Within 0.2 GPA Bandwidth")
```

```{r combinehs, fig.align='center', fig.cap="Distribution of Highschool Grade Within Certain GPA Bandwidth", fig.width=10, fig.height=5, echo= FALSE, message=FALSE, warning=FALSE}
# Referenced the following  http://www.sthda.com/english/articles/32-r-graphics-essentials/126-combine-multiple-ggplots-in-one-graph/
# combine figure into 1
ggarrange(hs_06,hs_02, ncol = 2, nrow = 1, common.legend = TRUE, legend = "bottom")
```

@@

## Test on Regression Discontinuity Design Assumptions {#assumption}

@@

&nbsp; &nbsp; Since the academic probation cutoff rule is known, one may question that students barely below the probation cutoff would study harder to avoid being placed on academic probation. However, the survey conducted in Lindo's paper suggests that first year students are generally less familiar with university policies to know the GPA required to avoid academic probation. And Lindo also finds no evidence that students' grades are related to their knowledge about academic probation [@cite_original]. So first year students are not likely to "play with" the university academic probation cutoff to avoid academic probation. The following figure (Figure \@ref(fig:histcutoff)) shows the distribution of the GPA distance from probation cutoff of all students. There is a smooth increase on the frequency distribution of the student GPA until reaching students' average GPA (GPA distance from probation around 1) and there is no obvious jump on either side of the probation cutoff. So students on either side of the probation cutoff is balanced as suggested in Data Section \@ref(bandwidth).

@@

ggplot(clean_for_summary, aes(x = dist_from_cut))+
  geom_histogram(aes(y  = stat(count)), binwidth = 0.125, color = "black", fill = "#33CCCC", alpha = 0.8)+
  geom_vline(xintercept = 0, linetype = "dotted")+
  theme_minimal()+
  theme(plot.title = element_text(hjust = 0.5)) +
  ylab("Frequency Headcount")+
  xlab("Year 1 GPA minus Probation Cutoff")+
  ggtitle("Distribution of Year 1 GPA relative to Probation Cutoff for All Students")

@@

&nbsp; &nbsp; By placing students on academic probation, the university informs these students an urge to improve their academic performance to prevent further escalating penalties such as suspension. According to $\mbox{B}{\acute{e}}\mbox{nabou}$ and Tirole's model of agents' responses to a performance standard [@cite_model_background], the probation policy as a negative incentive may result in the following outcomes:

@@

$$Y_{i} = \beta_{0}\mbox{Proation}_{i} + f(\mbox{Cutoff}\:\mbox{Distance}_{i}) + \alpha_{1}\mbox{Gender}_{i}+\alpha_{2}\mbox{HighSchool}_{i}+\alpha_{3}\mbox{Language}_{i} +\epsilon_{i}$$

@@

* $\mbox{Proation}_{i}$ is an indicator variable to show whether the student $i$ is placed on academic probation or not in the first year. And its coefficient $\beta_{0}$ is the estimate of impact on first year students after being placed on academic probation that we are primarily interested in.

* $f(\mbox{Cutoff}\:\mbox{Distance}_{i})$ is the function to represent the student $i$'s first year GPA distance from the cutoff. Since the first order polynomial is used during the power analysis to determine required student samples (Data Section \@ref(bandwidth)), it will also be applied in building the sharp RDD model^[Adjusting the order of distance from cutoff function does not affect the results. Lindo's original paper also used first order polynomial on distance from the cutoff to estimate probation outcomes].

* $\mbox{Gender}$ is an indicator variable to show whether the student $i$ is male or female. Its coefficient $\alpha_{1}$ is the estimate of gender difference in response to academic probation.

* $\mbox{HighSchool}_{i}$ is an indicator variable to show whether the student $i$ has a satisfactory highschool performance (Highschool Grade Percentile $>$ 50) or a relatively poor highschool performance (Highschool Grade Percentile $\leq$ 50). Its coefficient $\alpha_{2}$ is the estimate of highschool performance difference in response to academic probation.

* $\mbox{Language}_{i}$ is an indicator variable to show whether the student $i$'s first language is English or other language. Its coefficient $\alpha_{3}$ is the estimate of first language difference in response to academic probation.

@@

&nbsp; &nbsp; As mentioned in Model Section \@ref(model), logistic regression is used to estimate the probation impact on probability of student dropout.
$$Prob(\mbox{Dropout}_{i}) = \beta_{0}\mbox{Proation}_{i} + f(\mbox{Cutoff}\:\mbox{Distance}_{i}) + \alpha_{1}\mbox{Gender}_{i}+\alpha_{2}\mbox{HighSchool}_{i}+\alpha_{3}\mbox{Language}_{i} +\epsilon_{i}$$

@@

&nbsp; &nbsp; As mentioned in Model Section \@ref(model), linear regression is used to estimate the probation impact on student second year annual GPA.
$$\mbox{GPAYear2}_{i} = \beta_{0}\mbox{Proation}_{i} + f(\mbox{Cutoff}\:\mbox{Distance}_{i}) + \alpha_{1}\mbox{Gender}_{i}+\alpha_{2}\mbox{HighSchool}_{i}+\alpha_{3}\mbox{Language}_{i} +\epsilon_{i}$$

@@

&nbsp; &nbsp; Within 0.2 GPA bandwidth, there is significant difference on student second year GPA across gender and highschool performance. Compared to female students, male students earn 0.120 less annual GPA in the second year, holding other baseline characteristics constant. Compared to students with relative poor highschool academic performance, students with satisfactory highschool academic performance earn 0.099 more annual GPA in the second year, holding other baseline characteristics constant. No other significant difference is found on student first language.
```{r, include=FALSE}
# Linear Regression to estimate the impact of probation on student Year 2 GPA

@@

&nbsp; &nbsp; As mentioned in Model Section \@ref(model), logistic regression is used to estimate the probation impact on probability of student graduating by Year 4.
$$Prob(\mbox{Graduation}_{i}) = \beta_{0}\mbox{Proation}_{i} + f(\mbox{Cutoff}\:\mbox{Distance}_{i}) + \alpha_{1}\mbox{Gender}_{i}+\alpha_{2}\mbox{HighSchool}_{i}+\alpha_{3}\mbox{Language}_{i} +\epsilon_{i}$$
&nbsp; &nbsp; The model baseline characteristics are the same as mentioned in dropout Section \@ref(baseline). As shown in the following table (Table \@ref(tab:gradsummary)), within 0.2 GPA bandwidth, compared to students on academic probation, students in good standing have $26.8\%$ increase in the odds of graduating by Year 4, while holding other baseline characteristics constant. **However**, the p-value of 0.161 suggests that the graduation does not significantly differ between good standing students and probation students, holding other baseline characteristics constant. And this difference could also not be clearly detected around the probation cutoff as shown(Figure \@ref(fig:gradfig)). Within 0.3 GPA and 0.6 GPA bandwidth, the results also do not suggest difference on graduation between good standing students and probation students at 0.05 significance level, holding other baseline characteristics constant.

@@

&nbsp; &nbsp; According to the student summary statistics (Table \@ref(tab:lindo)) and t-test conducted on students highschool performance (Table \@ref(tab:ttesthslin)) and credit attempted in the first year (Table \@ref(tab:ttestcrlin)) in Data Section \@ref(bandwidth), within the bandwidth of 0.6 GPA, students below the cutoff generally hold weaker academic potentials than those above the cutoff. And these academic differences also serves as key factors affecting student outcomes.

@@

## Causal Inference, Internal Validity and External Validity

### Causal Inference and Internal Validity

&nbsp; &nbsp; So does placing first year students on academic probation affect student outcomes? Instead of randomly assigning students either probation or good standing status as a randomized controlled trial, this study uses regression discontinuity design to estimate the probation impact through observed student outcomes around the probation cutoff. And one may wonder if such estimation is able to infer that academic probation causes changes on student outcomes. Actually, as students get closer to the cutoff, they share such similar characteristics that probation estimation could be as good as randomized controlled trial. Besides, previous meta-analysis on RDD has also suggested its high internal validity to provide an accurate probation estimate within the study settings [@cite_internal].

&nbsp; &nbsp; By applying a narrower bandwidth of 0.2 GPA, student characteristics become more balanced (Table \@ref(tab:mycutoff)). Although the t-test still shows a difference on highschool performance between treatment and comparison students (Table \@ref(tab:ttesthslin)), study results in Result Section \@ref(result) has already suggested that the impact of academic probation on student dropout, Year 2 annual GPA and graduation is not "significant" under relevant models. In fact, further restricting the bandwidth to 0.1 GPA also does not show "significant" probation impact on student outcomes.

&nbsp; &nbsp; However, as mentioned in Data Section \@ref(bandwidth), narrower GPA bandwidth will include less student samples, lowering the power of the model. With insufficient power, p-value obtained from the model is more likely to happen by chance instead of the probation cutoff rule, so its interpretation may also be in vain.

&nbsp; &nbsp; Therefore, from model results of 0.2 GPA bandwidth, in this Canadian university, there is "no evidence" to show that the academic probation policy causes:

* Increase in chances of student dropout

* Increase in second year annual GPA of remaining students

* Decrease in chances of graduation by Year 4

&nbsp; &nbsp; While there is "evidence" to show that after being placed on academic probation in the first year:

* Students whose first language are not English are less likely to dropout school

* Male students earn less second year annual GPA, and are less likely to graduate by the fourth year than female students

* Students with high highschool grades earn more annual GPA than whose with low highschool grades

### Campus Variation and External Validity

&nbsp; &nbsp; As the data used in this study is obtained from a large Canadian university, can the study results generalize to other universities that also have academic probation or related policies? Not likely.

```{r hsgbox, fig.align='center', fig.cap="Distribution of Student Highschool Grades by Campus", fig.width=5, fig.height=2.5, echo= FALSE, message=FALSE, warning=FALSE}
# boxplot to show difference in highschool grade across campus
ggplot(clean_for_summary, aes(x = study_campus, y = hsgrade_pct, color = study_campus))+
  geom_boxplot()+
  theme_minimal()+
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5))+
  labs(x = "Study Campus", y = "Highschool Grade Percentile",
       title = "Distribution of Student Highschool Grades by Campus")
```

&nbsp; &nbsp; In Institutional Background Section of Lindo's paper, they showed a lower acceptance rate (about $55\%$) in Campus 1 and a higher acceptance rate (about $77\%$) in Campus 2 and 3. The above boxplot (Figure \@ref(fig:hsgbox)) also suggests highschool grade difference across campus. Besides, Campus 1 resembles a large US state college and Campus 2 and 3 have more part time and commuter students [@cite_original]. Despite the university share same probation policy, students may still react differently towards academic probation under different campus environment. So I add a campus variable into the model constructed in Model Section \@ref(model) to examine if there are campus variations towards academic probation:
$$Y_{i} = \beta_{0}\mbox{Proation}_{i} + f(\mbox{Cutoff}\:\mbox{Distance}_{i}) + \alpha_{1}\mbox{Gender}_{i}+\alpha_{2}\mbox{HighSchool}_{i}+\alpha_{3}\mbox{Language}_{i} + \alpha_{4}\mbox{Campus}_{i} + \epsilon_{i}$$
&nbsp; &nbsp; From model results of student outcomes as shown below (Table \@ref(tab:campus)), students in different campus do react differently towards academic probation: Compared to students in Campus 1: (1) Students in Campus 3 are less likely to leave the university, and (2) Students in Campus 3 earn more annual GPA in the second year and are more likely to graduate by Year 4, holding other baseline characteristics constant.

&nbsp; &nbsp; As a result, campus variation towards academic probation demonstrates limited external validity of this study. Students are not likely to react similarly towards academic probation across university or college, whose demographics and general academic standards are usually different. More importantly, each university or college generally holds distinct policies towards academic probation, and it is not appropriate to use a specific probation policy in one university to infer academic probation impact on other universities solely from this study.

```{r, include=FALSE}
# Add Campus variable into the model
gpa_camp2 <- lm(GPA_year2 ~ status + dist_from_cut + gender + first_language + highHS + study_campus, data = clean_cut)
#summary(gpa_camp2)
dropout_camp2 <- glm(left_school ~ status + dist_from_cut + gender + first_language + highHS + study_campus, family = "binomial", data = clean_cut)
#summary(dropout_camp2)
grad_camp2 <- glm(gradin4 ~ status + dist_from_cut + gender + first_language + highHS + study_campus  ,family = "binomial", data = clean_cut)
#summary(grad_camp2)
```
